Revisiting~the~Classics: Online~RL~in~the~Programmable~Dataplane

### 2021-06-09 paper/paper.tex
Two More Paras

\setlength{\textfloatsep}{4.0pt plus 2.0pt minus 4.0pt}
\setlength{\floatsep}{4.0pt plus 1.0pt minus 2.0pt}

\setlength{\dbltextfloatsep}{4.0pt plus 2.0pt minus 4.0pt}
\setlength{\dblfloatsep}{4.0pt plus 1.0pt minus 2.0pt}

@@

\section{Preliminaries: RL for in-network computation}\label{sec:motivation}
We discuss recent trends in programmable switch hardware.
By combining this with insights from the ML/RL communities (past and present), we discuss why in-NIC RL is best-placed to interact with the network, and how classical RL methods and quantisation ensure computational feasibility.

@@

				\draw[dotted, ->, bend left = 30] (state) -- (agent) node[midway, above] {$t_1$};
				\draw[->] (agent) -- (action) node[midway, right] {$t_2$};
				\draw[dotted, ->] (action) -- (state') node[midway, below] {$t_3$};
		\end{tikzpicture}}
		\caption{Reality: costs of measurement and action lead to \emph{state drift}---over a time delay $t_1+t_2+t_3$, inaction transforms state $S$ into $S'$.}
	\end{subfigure}
	\caption{Asynchronous RL delays and state slippage.\label{fig:state-slip}}
\end{figure}

@@

For instance, a policy with $m$ sets of tiles (each having its own set of input dimensions), each containing $n$ overlapping tilings then creates $m \times n$ separate tasks.\footnote{Retrieving each individual action's value can be considered a work item (giving $m \times n \times a$ tasks for $a$ actions). However, this requires tile coding to be repeated $a$ times per tile, which is likely extremely wasteful outside of heavy parallelism.}
As an example, a real-world policy size which we examine later (\cref{sec:experimental-setup}) with one bias tile and \num{16} sets containing \num{8} tilings creates \num{129} work items.
Computing a tile hit requires one division per dimension, which is an expensive addition to the ALU operations we require from a platform.
However, if the width of tiles are fixed at powers-of-two, then these may be replaced with right bitshifts.
%As we shall see shortly, this aligns with our choice of numerical representation.

@@

\Cref{fig:netro-arch,fig:single-and-parallel} outline our design and implementation on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec \emph{can and should be used} to drive asynchronous environmental control.
We explain relevant NFP architectural details in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently (\cref{fig:single-and-parallel:single}) or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits (\cref{sec:action-and-update-computation}).
We open-source our firmware design and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
Moving beyond the Netronome platform, we describe how our architectural choices may be adapted and improved upon by more bespoke hardware or FPGA-based deployment.

@@

To protect traffic throughput and allow for effective deployment in as many environments as possible, \approachshort{} places RL execution on-chip, \emph{but off the main packet path}, communicating and running parallel to the main P4 dataplane.
As shown in \cref{fig:netro-arch}, this asynchrony allows coexistence with P4 programs, and imposes minimal impact on carried traffic for both bump-in-the-wire deployments and at end-points.
For instance, in the default deployment of a P4 packet processing pipeline on Netronome NFP chips several of these cores go unused (as does spare area on an FPGA design), making this paradigm possible.

@@

Two main firmware models govern how the computation-intensive parts of these tasks (tile-coding, action preference list construction, policy updates) are carried out:
\begin{description}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each performs its work sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, greatly limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate on processing state vectors, minimising latency. \emph{Minions} have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{description}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks for updating the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile-coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput, as in many networked systems, have different effects upon RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem, or queue depth when trying to enforce packet pacing).

@@

\subsection{Agent-Environment Communication}\label{sec:agent-environment-communication}
\approachshort{} uses \emph{multiple-producer/multiple-consumer} (MPMC) messaging channels to communicate with other computational elements; be they P4 programs on the packet path, or additional on-chip analysis and control modules.
Through these channels the system \emph{pushes} state vectors, reward measures, and control/setup packets as its inputs, and \emph{pulls} a stream of state-action pairs as its outputs.
This allows decisions to be made asynchronously---preventing packet stalling---and allowing many parallel decision-making agents to be used if desired.
The key insight of this mechanism is that on-chip reward/state signals enjoy first-class support in the same manner as packets from the P4 dataplane, allowing agents to act on environmental signals from other on-NIC/chip asynchronous processes or the controller.
As such, \approachshort{} can receive from P4 \texttt{extern}s or another dedicated off-path flow state measurement application.

Our implementation uses platform-specific IPC (\emph{EMEM ring buffers}) with hardware signalling mechanisms on work arrival to achieve this.
As PDP hardware lacks dynamic memory allocation, P4 pipeline threads request buffers for packet payloads using a shared freelist to enable state/configuration/policy data handover.
Packet headers are extracted and parsed using the tooling autogenerated by the P4 pipeline.
We found that this costs a median \SIrange{126}{140}{\nano\second} communication time (local--remote), comparable to message channels in the Rust and Go languages on commodity hardware.

@@

\subsection{Intra-Agent Communication}\label{sec:intra-agent-communication}
Even with parallel problems such as \emph{ParSa}, optimising for latency requires meticulous care in how work is passed out and aggregated.
This is truer still when moving from the moderately fine-grained control of classical RL methods ($\sim$\SI{1}{\milli\second}) to its logical limit (tens of \si{\micro\second}).
Ordinarily, the marshalling of requests, responses, and safe shared data access can incur significant extra overheads.
On-chip execution and the nature of action preference computation allow us to use lockless atomic aggregation, removing the overheads of explicit messaging/packetisation.
Moreover, physically adjacent functional units/cores often have special-purpose shared registers or share a small fast cache to accelerate communication.

@@

Our implementation exploits the locality of cores in the NFP architecture, which can be used on other platforms like NetFPGA.
Policy compute/update/configure tasks are passed between cores using direct \emph{next neighbour} registers, signalling all child threads in response.
Such on-chip signals cost just $\sim$\SI{20}{\nano\second} per relayed message.
Each core performs atomic adds to a shared preference list and an acknowledgement counter checked by the master thread, implementing the wait-free \emph{ParSa} algorithm we introduce (\cref{alg:parsa}).
This is essential for aggregation compared to the use of bounded message buffers, which caused significant head-of-line blocking.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and algorithmic learning parameters ($\alpha, \gamma, \epsilon$, related falloffs) to be changed at runtime using at most two control packets.
For instance, design changes are useful at the end of learning (moving from online to offline), or when trying to train a policy for another problem afterwards from the same vantage point.
Parameter changes are useful when an online agent must become more (or less) adaptive to new data (i.e., after a changepoint in traffic is detected).
This extends to policy data, which may be imported from an offline pre-trained model via such packets and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, work allocation strategy, and maximum policy/tiling/state sizes---these govern core operation or allocated memory.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing more complex problems to be modelled using more dimensions or fine-grained tiles.

In our implementation, configuration packets are carried over UDP and signalled to the P4 parser using a reserved (pool 2~\parencite{rfc2474}) DSCP value as used by \textcite{DBLP:conf/isca/LiLYCSH19}.
While we primarily use this to automate parser generation, this also allows for configuration packets to be received from only trusted hosts (over the dataplane if needed) via P4 rules.
Our control packet generation library (and evaluation frameworks which build upon it) are written in Rust.

@@

\subsection{Work Allocation}\label{sec:work-allocation}
Due to the lack of dynamic memory allocation on PDP hardware, and to simplify action-value lookups, policies cannot be stored sparsely.
As a consequence, tiling space requirements scale exponentially with dimension count, and so higher-dimension tilings must be placed in larger (and slower) memory regions.
As a result, in many architectures policy parameters are split across such regions, giving different access and compute costs to different tasks.

We use a simple first-fit work placement algorithm run in \approachshort{}, placing the largest work item into the least loaded thread of the least loaded core (as the NFP exposes hardware threads).
Each work item is a separate \emph{tiling} over a number of dimensions, noting that many such tilings may be grouped into a set (using different offsets for smoother coverage of the state space).
The cost of any individual work item (according to its dimension count and memory location) was empirically measured offline, and we weigh the total cost per core to account for the number of minion threads available.
This weighting specifically accounts for the presence of the controller thread on the first core.
Work allocation and any cached per-item data are recomputed when any configuration is installed or changed.
Naturally, for $n$ tilings/tasks and $m$ threads this procedure is $\mathcal{O}{\left(n\log{m}\right)}$: two find/update min operations into binary heaps per task, storing $m/8$ and $\le8$ costs respectively.

@@

As is expected of parallel algorithms, efficient division of work, communication, and aggregation add per-task overheads in both the serial and parallel portions.
Even in wait-free algorithms, this requires a minimum number of workers to improve upon the latency bounds of a serial approach.
We investigate the exact worker count requirements imposed to break even or improve upon single-threaded execution in the sequel.

\subsection{Targeting other device classes}
While specifics of this design cater to Netronome devices, many aspects have clear analogues in NICs of similar form-factor, such as the NetFPGA SUME platform.
For instance, other cores can be introduced by designing additional off-path functional units,  like the floating-point adders used by \textcite{DBLP:conf/isca/LiLYCSH19}.
Rather than general-purpose IPC, \approachshort{} would be hard-wired between internal workers and to the custom actions that interface with it, using the P4$\rightarrow$NetFPGA toolchain to offer granular control over packet and flow selection.
%Further optimisations arise from this model: a NetFPGA design would be able to replicate and optimise further on this concept of dedicated, low-latency communication between functional units.
As our later results show, using a bit depth below the native word size forces the compiler to emit excess code to extract and re-align preference values, reducing overall efficiency.
%This reduces overall efficiency even though considerably fewer reads are made.
We expect that a more bespoke hardware/FPGA design would allow this to match the desired bit depth and enable SIMD-like optimisations we discuss without creative and costly bit-stuffing.

Bringing \approachshort{} to high-port density devices such as Tofino-powered switches is more difficult.
As discussed in \cref{sec:motivation}, Tofino ASICs map closely to the P4 PSA, meaning that there are no spare asynchronous general-purpose compute units to place this work on.
Taking inspiration from real-time programming techniques, a potential solution is to divide RL actions across several received packets (i.e., iteratively computing a portion of the action preference list each time) until any further work would delay outbound transmission.
This, however, would introduce new issues surrounding concurrent accesses, work splitting, and altered timescales for learning: we leave their treatment and examination to future work.

\section{Evaluation}\label{sec:evaluation}
We investigate the performance of \approachshort{} compared to classical RL techniques executed on commodity host machines, with \Coopfw{} offering a \SIrange{15}{21}{$\times$} speedup in median--\num{99.99}\nthscript{th} state-action latency and \SI{9.9}{$\times$} greater online learning throughput.
Crucially, we find that in-NIC execution offers tight tail latency bounds compared to host-based approaches.
We report on how \approachshort{} scales to fit available device resources as cores/threads are added, noting that both \Indfw{} and \Coopfw{} designs outperform commodity hosts using just one core in latency and online throughput.
Furthermore, \Indfw{} provides higher per-core offline throughput than host-based approaches, even though our measured hosts exhibit higher clock speeds.
Finally, we show that \approachshort{} has minimal impact on dataplane cross-traffic carried by its parent device.

@@

We run \approachshort{} on a\num{4}-ME island of the NFP-6480, totalling \num{32} contexts.
This is the largest cluster of cores which is not in use by a P4 pipeline.
Where feasible, we test \SIlist{32;16;8}{\bit} quantised arithmetic.
All \approachshort{} timing measurements were repeated over \num{10000} state packets (preceded by \num{1000} warmup packets), retrieving item processing times over PCIe via the controller machine, from which throughput was derived.
Host throughput and latency measures were observed over \num{10} trials of \SI{10}{\second} (with \SI{5}{\second} warmup/cooldown times).
We differ this from the NFP as hosts need to run numpy-based agents in parallel via separate processes; this also allows us to investigate the effects of oversubscription on host latencies.

Policy sizes are set to those matching a real-world DDoS control application~\parencite{DBLP:journals/tnsm/SimpsonRP20}: 20-dimension state vectors, a bias tile and 16 full tiling sets (\numproduct{7x1}-dim, \numproduct{8x2}-dim, \numproduct{1x4}-dim), 8 tilings per dimension set, 6 tiles per dimension, choosing between 10 actions.
For context, such input state would contain various aspects of per-flow state (e.g., IATs, up/down rate) which are combined with other state such as the last action taken (2-dim tilings) and loads along the ingress-egress path (4-dim).
In \Coopfw{} for instance, this creates \num{129} work items across \num{31} workers.
Although the model application's more successful \emph{SPF} agent design uses 3 actions, we choose a larger action set to investigate the performance of more complex agents.

@@

\fakepara{Raw inference and learning performance}
We compare how long it takes for \approachshort{} to compute actions and update its policy per state vector received, and report on the observed throughput of our designs against floating point (numpy-based) implementations of Sarsa on a commodity host machine.
This allows us to demonstrate the performance differences between the \indfw{} and \coopfw{} configurations, particularly in how \indfw's required policy locks impact throughput.
%We also report on the derived throughput measures accounting for the locking behaviour of each.
We compare online learning performance (input states produce an output, and then update the policy) with offline (input states \emph{only} produce an output) in these cases.
Online performance marks the number of decisions that can be made per second (and associated latency) when training a policy.
Offline performance is crucial for pushing a trained, known-good policy to agents in the network with an expected higher raw decision throughput.
State-action latency is a shared property of both cases, with the main impact on throughput arising from the update step.

Building on this, we vary the amount of allocated worker threads to show how \approachshort{} can scale to fit available compute resources on a device.
This is an important aspect for (automated) administration and allocation of compute resources in an intelligent dataplane---particularly when cohabiting with other dataplane programs---and has effects on ahead-of-time work scheduling which we examine later.
This also demonstrates the number of cores needed to achieve a given latency/throughput bound on a policy of representative complexity.
Moreover, to demonstrate how these costs vary as policy complexity increases, we vary the number of total dimensions included in the tiling.

@@

\newlength{\resultplotwidth}
\setlength{\resultplotwidth}{0.9\linewidth}

@@

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves with additional cores, on max size tasks (\num{129} work items). This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-latency}
	}
	\caption{\Coopfw{}'s action latencies similarly improve with additional cores. This requires \num{3} workers (4 total contexts) to improve upon the state-action latency of single-threaded in-NIC RL.\label{fig:vary-core-latency}}
\end{figure}

@@

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-work}
	}
	\caption{\Coopfw{} fully processes updates faster than \Indfw{}---thus has higher online performance---on almost all policy sizes. Lower bit depths are more effective on simpler policies.\label{fig:vary-work}}
\end{figure}

>>>>

\setlength{\textfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
\setlength{\floatsep}{3.0pt plus 0.0pt minus 2.0pt}

\setlength{\dbltextfloatsep}{3.0pt plus 0.0pt minus 3.0pt}
\setlength{\dblfloatsep}{3.0pt plus 0.0pt minus 2.0pt}

@@

\section{Preliminaries: RL for in-network computation}\label{sec:motivation}
We discuss recent trends in programmable switch hardware.
By combining this with insights from the ML/RL communities (past and present), we discuss why in-NIC RL is needed and best-placed to interact with the network, and how classical RL methods and quantisation make this computationally feasible.

@@

				\draw[dotted, ->, bend left = 30] (state) -- (agent) node[midway, above] {$t_1$};
				\draw[->] (agent) -- (action) node[midway, right] {$t_2$};
				\draw[dotted, ->] (action) -- (state') node[midway, below] {$t_3$};
		\end{tikzpicture}}
		\caption{Reality: measurement and action costs lead to \emph{state drift}---over a delay $t_1+t_2+t_3$, inaction transforms state $S$ into $S'$.}
	\end{subfigure}
	\caption{Asynchronous RL delays and state slippage.\label{fig:state-slip}}
\end{figure}

@@

For instance, a policy with $m$ sets of tiles (each having its own set of input dimensions), each containing $n$ overlapping tilings then creates $m \times n$ separate tasks.\footnote{Retrieving each individual action's value can be considered a work item (giving $m \times n \times a$ tasks for $a$ actions). However, this requires tile coding to be repeated $a$ times per tile, which is likely extremely wasteful outside of heavy parallelism.}
As an example, a real-world policy size which we examine later (\cref{sec:experimental-setup}) with one bias tile and \num{16} sets containing \num{8} tilings creates \num{129} work items.
Computing a tile hit requires one division per dimension, which is an expensive ALU operation to require.
However, if tile widths are fixed at powers-of-two, then these may be replaced with right bitshifts.
%As we shall see shortly, this aligns with our choice of numerical representation.

@@

\Cref{fig:netro-arch,fig:single-and-parallel} outline our design and implementation on Netronome SmartNIC hardware in pursuit of this goal: unused device resources beyond the P4-PSA spec \emph{can and should be used} to drive asynchronous environmental control.
We explain relevant NFP architectural details in \cref{sec:netronome-platform-fundamentals}.
\approachshort{} communicates with the packet pipeline of a P4 dataplane via \texttt{extern} plugins using \inring{} (state, configuration) and \outring{} (action) messages (\cref{fig:netro-arch}).
Internally, \approachshort{} either has all its cores act independently (\cref{fig:single-and-parallel:single}) or cooperate to solve each task (\cref{fig:single-and-parallel:parallel})---with different latency-throughput benefits (\cref{sec:action-and-update-computation}).
We open-source our firmware and control programs for the benefit of the community.\footnote{\textanon{\url{https://github.com/FelixMcFelix/pdp-rl-paper}}{Link redacted for double-blind submission.}}
Moving beyond the Netronome platform, we describe how our architecture may be adapted and improved on by bespoke hardware or FPGA-based deployment.

@@

To protect traffic throughput and allow effective deployment in as many environments as possible, \approachshort{} places RL execution on-chip, \emph{but off the main packet path}, communicating and running parallel to the main P4 dataplane.
As shown in \cref{fig:netro-arch}, this asynchrony allows coexistence with P4 programs, and imposes minimal impact on carried traffic for both bump-in-the-wire deployments and at end-points.
For instance, in the default deployment of a P4 packet processing pipeline on Netronome NFP chips several cores go unused (as does spare area on an FPGA design), making this paradigm possible.

@@

Two firmware models govern how the compute-heavy parts of these tasks (action selection, policy updates) are carried out:
\begin{description}
	\item[\Indfw{} (\cref{fig:single-and-parallel:single})] Separate threads listen for new states, and each performs its work sequentially. Computing an action list requires a \emph{read lock} on the policy. If an update occurs, the core requests a \emph{write lock} before updating, greatly limiting online throughput. \emph{Tile lists} are stored for update computation.
	\item[\Coopfw{} (\cref{fig:single-and-parallel:parallel,alg:parsa})] Threads cooperate on processing state vectors, minimising latency. \emph{Minions} have a fixed list of work items, while a \emph{controller} thread sends compute/update commands before awaiting worker completion. Work items are disjoint, requiring no policy locks. \emph{State vectors} are stored for update computation.
\end{description}
Each offers a different point of optimisation; if updates are disabled, then the \indfw{} model can maximise throughput, while the \coopfw{} model is designed to minimise decision latency and needs no locks for updating the policy (increasing \emph{online learning} throughput).
These correspond to only executing a trained policy and actively (re-)training a policy, respectively.
%ParSa is described here at a high level, as an algorithm suited for \emph{any multiprocessor environment}.
%We omit the simple logic for $\epsilon$-greedy action selection (which we implement), and note that modification to other single-step algorithms such as \emph{Q-learning} would be trivial.
%We detail our communication primitives in \cref{sec:intra-agent-communication}, and our work scheduling strategy in \cref{sec:work-allocation}, eliding the details of tile-coding as they are well-understood.
%In general it is also possible for the \emph{Ctl} task to act as an additional \emph{Minion} in its parallel sections; we were limited here by code store requirements on the NFP.
Latency and throughput, as in many networked systems, have different effects upon RL agents according to their design and target problem.
Higher RL throughput is a necessity for per-flow or per-packet applications, which can require high decision-per-second rates even after combining state measurements received from the environment, such as flow control in DDoS prevention.
Equally, lower latency affords an agent finer-grained control and learning of a problem, being able to react sooner to new information (e.g., device state in a routing optimisation problem, or queue depth when trying to enforce packet pacing).

@@

\subsection{Agent-Environment Communication}\label{sec:agent-environment-communication}
\approachshort{} uses \emph{multiple-producer/multiple-consumer} (MPMC) messaging channels to communicate with other elements; be they P4 programs on the packet path, or other on-chip analysis and control modules.
Through these channels a system \emph{pushes} state vectors, reward measures, and setup packets as inputs, and \emph{pulls} a stream of state-action pairs as outputs.
This allows decisions to be made asynchronously---preventing packet stalling---and allowing many RL agents to be used if desired.
The key insight of this mechanism is that on-chip reward/state signals enjoy first-class support in the same manner as packets from the P4 dataplane, allowing agents to act on environmental signals from other on-NIC/chip asynchronous processes or the controller.
As such, \approachshort{} can receive from P4 \texttt{extern}s or another dedicated off-path flow state measurement application.

Our implementation uses platform-specific IPC (\emph{EMEM ring buffers}) with hardware signalling on work arrival to achieve this.
As PDP hardware lacks dynamic memory allocation, P4 pipeline threads request buffers for packet payloads using a shared freelist to enable state/configuration/policy data handover.
Packet headers are extracted and parsed using the tooling autogenerated by the P4 pipeline.
We found that this costs a median \SIrange{126}{140}{\nano\second} communication time (local--remote), comparable to message channels in the Rust and Go languages on commodity hardware.

@@

\subsection{Intra-Agent Communication}\label{sec:intra-agent-communication}
Even with parallel problems such as \emph{ParSa}, optimising for latency requires meticulous care in how work is passed out and aggregated.
This is truer still when moving from the moderately-grained control of classical RL ($\sim$\SI{1}{\milli\second}) to its logical limit (tens of \si{\micro\second}).
Ordinarily, the marshalling of requests, responses, and shared data access can incur significant overheads.
On-chip execution and the nature of action preference computation allow us to use lockless atomic aggregation, removing the overheads of explicit messaging/packetisation.
Moreover, adjacent functional units/cores often have special-purpose shared registers or share a small fast cache to accelerate communication.

@@

Our implementation exploits the locality of cores in the NFP, which can be used on other platforms like NetFPGA.
Policy compute/update/configure tasks are passed between cores using direct \emph{next neighbour} registers, signalling all child threads in response.
Such on-chip signals cost just $\sim$\SI{20}{\nano\second} per relayed message.
Each core performs atomic adds to a shared preference list and an acknowledgement counter checked by the master thread, implementing our wait-free \emph{ParSa} algorithm (\cref{alg:parsa}).
This is essential for aggregation compared to the use of bounded message buffers, which caused significant head-of-line blocking.

@@

\subsection{Reconfigurability}
\approachshort{} allows policy design and algorithmic learning parameters to be changed at runtime using at most two control packets.
For instance, design changes are useful at the end of learning (moving from online to offline), or when trying to train a new policy for another problem from the same vantage point.
Parameter changes are useful when an online agent must become more (or less) adaptive to new data (i.e., after detecting a changepoint in traffic).
This extends to policy data, which may be imported from a pre-trained model via such packets and exported via PCIe to the host machine.
Some aspects must be chosen at compile time; bit depth, \Coopfw/\Indfw, and maximum policy/tiling/state sizes---these govern core operation or allocated memory.
Choosing a bit depth of \qtylist[list-pair-separator = { or }]{16;8}{\bit} halves/quarters policy memory costs, allowing more complex problems to be modelled using more dimensions or fine-grained tiles.

In our implementation, configuration packets are carried over UDP and signalled to the P4 parser using a reserved (pool 2~\parencite{rfc2474}) DSCP value as used by \textcite{DBLP:conf/isca/LiLYCSH19}.
While this mainly automates parser generation, it also allows for configuration to be received from only trusted hosts (over the dataplane if needed) via P4 rules.
Our control packet generation library and evaluation frameworks which build upon it are written in Rust.

@@

\subsection{Work Allocation}\label{sec:work-allocation}
Due to the lack of dynamic memory allocation on PDP hardware, and to simplify value lookups, policies cannot be stored sparsely.
As a consequence, tiling space requirements scale exponentially with dimension count, and so higher-dimension tilings must be placed in larger (and slower) memory regions.
As a result, in many architectures policy parameters are split across such regions, giving different access and compute costs to different tasks.

We use a simple first-fit work placement algorithm run in \approachshort{}, placing the largest work item into the least loaded thread of the least loaded core (as the NFP uses hardware threads).
Each work item is a separate \emph{tiling} over a list of dimensions, noting that many such tilings may be grouped into a set (using different offsets for smoother coverage of the state space).
The cost of any work item (from its dimension count and memory location) was empirically measured offline, and we weigh the total cost per core based on the number of minion threads available.
This weighting specifically accounts for the controller thread on the first core.
Work allocation and cached per-item data are recomputed when policy configuration is installed or changed.
Naturally, for $n$ tilings and $m$ threads this procedure is $\mathcal{O}{\left(n\log{m}\right)}$: two find/update min operations into binary heaps per tiling, storing $m/8$ and $\le8$ costs respectively.

@@

%As is expected of parallel algorithms, efficient division of work, communication, and aggregation add per-task overheads in both the serial and parallel portions.
Efficient division of work, communication, and aggregation add per-task overheads in both the serial and parallel portions.
Even in wait-free algorithms, this requires a minimum number of workers to improve upon the latency bounds of a serial approach.
We investigate the exact worker count requirements imposed to break even or improve upon single-threaded execution in the sequel.

\subsection{Targeting other device classes}
While this design caters to Netronome devices, many aspects have clear analogues in other SmartNICs, such as the NetFPGA SUME.
For instance, additional off-path functional units can serve as workers, like the floating-point adders used by \textcite{DBLP:conf/isca/LiLYCSH19}.
Rather than general-purpose IPC, \approachshort{} would be hard-wired between internal workers and to the P4$\rightarrow$NetFPGA~\parencite{DBLP:conf/fpga/IbanezBMZ19} actions that interface with it.
%Further optimisations arise from this model: a NetFPGA design would be able to replicate and optimise further on this concept of dedicated, low-latency communication between functional units.
As we show later, using a bit depth below the native word size reduces overall efficiency.
%This reduces overall efficiency even though considerably fewer reads are made.
We expect an FPGA design would allow this to match the desired bit depth and enable the SIMD-like optimisations we discuss without creative and costly bit-stuffing.

Bringing \approachshort{} to high-port density devices such as Tofino switches is more difficult as they map closely to the P4-PSA, leaving no spare general-purpose compute units.
Taking inspiration from real-time programming, a potential solution is to divide RL processing across several packets (i.e., computing a portion of the action preference list each time) until any further work would delay outbound transmission.
This, however, would introduce new issues surrounding concurrent accesses, work splitting, and altered timescales for learning: we leave their treatment and examination to future work.

%While specifics of this design cater to Netronome devices, many aspects have clear analogues in NICs of similar form-factor, such as the NetFPGA SUME platform.
%For instance, other cores can be introduced by designing additional off-path functional units,  like the floating-point adders used by \textcite{DBLP:conf/isca/LiLYCSH19}.
%Rather than general-purpose IPC, \approachshort{} would be hard-wired between internal workers and to the custom actions that interface with it, using the P4$\rightarrow$NetFPGA toolchain to offer granular control over packet and flow selection.
%%Further optimisations arise from this model: a NetFPGA design would be able to replicate and optimise further on this concept of dedicated, low-latency communication between functional units.
%As our later results show, using a bit depth below the native word size forces the compiler to emit excess code to extract and re-align preference values, reducing overall efficiency.
%%This reduces overall efficiency even though considerably fewer reads are made.
%We expect that a more bespoke hardware/FPGA design would allow this to match the desired bit depth and enable SIMD-like optimisations we discuss without creative and costly bit-stuffing.
%
%Bringing \approachshort{} to high-port density devices such as Tofino-powered switches is more difficult.
%As discussed in \cref{sec:motivation}, Tofino ASICs map closely to the P4 PSA, meaning that there are no spare asynchronous general-purpose compute units to place this work on.
%Taking inspiration from real-time programming techniques, a potential solution is to divide RL actions across several received packets (i.e., iteratively computing a portion of the action preference list each time) until any further work would delay outbound transmission.
%This, however, would introduce new issues surrounding concurrent accesses, work splitting, and altered timescales for learning: we leave their treatment and examination to future work.

\section{Evaluation}\label{sec:evaluation}
We investigate the performance of \approachshort{} compared to classical RL techniques executed on commodity host machines, with \Coopfw{} offering a \SIrange{15}{21}{$\times$} speedup in median--\num{99.99}\nthscript{th} state-action latency and \SI{9.9}{$\times$} greater online learning throughput.
Crucially, in-NIC execution offers tight tail latency bounds compared to host-based approaches.
We report on how \approachshort{} scales as additional device resources are added, noting that both in-NIC designs outperform commodity hosts using just one core in latency and online throughput.
Furthermore, \Indfw{} provides higher per-core offline throughput than host-based approaches, even though our measured hosts exhibit higher clock speeds.
Finally, we show that \approachshort{} has minimal impact on dataplane cross-traffic carried by its parent device.

@@

We run \approachshort{} on a \num{4}-ME island of the NFP-6480, totalling \num{32} contexts.
This is the largest cluster of cores which is not in use by a P4 pipeline.
Where feasible, we test \SIlist{32;16;8}{\bit} quantised arithmetic.
All \approachshort{} timing measurements were repeated over \num{10000} state packets (preceded by \num{1000} warmup packets), retrieving item processing times over PCIe via the controller machine, from which throughput was derived.
Host throughput and latency measures were observed over \num{10} trials of \SI{10}{\second} (with \SI{5}{\second} warmup/cooldown times).
We differ this from the NFP as hosts need to run numpy-based agents in parallel via separate processes; this also allows us to investigate the effects of oversubscription on host latencies.

Policy sizes are set to those of a real-world DDoS control application~\parencite{DBLP:journals/tnsm/SimpsonRP20}: 20-dim state vectors, a bias tile and 16 full tiling sets (\numproduct{7x1}-dim, \numproduct{8x2}-dim, \numproduct{1x4}-dim), 8 tilings per set, 6 tiles per dimension, and 10 actions.
For context, such input state would contain various aspects of per-flow state (e.g., IATs, rates) which are combined with other state such as the last action taken (2-dim tilings) and loads along the ingress-egress path (4-dim).
In \Coopfw{}, this creates \num{129} work items across \num{31} workers.
Although the model application's more successful \emph{SPF} agent design uses 3 actions, we choose a larger action set to investigate the performance of more complex agents.

@@

\fakepara{Raw inference and learning performance}
We compare how long it takes for \approachshort{} to compute actions and update its policy per state vector received, and report on the observed throughput of our designs against floating point (numpy-based) implementations of Sarsa on a commodity host machine.
This allows us to demonstrate the performance differences between the \indfw{} and \coopfw{} configurations, particularly in how \indfw's (and hosts') required policy locks impact throughput.
%We also report on the derived throughput measures accounting for the locking behaviour of each.
We compare online learning performance (input states produce an output, and then update the policy) with offline (input states \emph{only} produce an output) in these cases.
Online performance marks the number of decisions that can be made per second (and associated latency) when training a policy.
Offline performance is crucial for pushing a trained, known-good policy to agents in the network with an expected higher raw decision throughput.
State-action latency is a shared property of both cases, with the main impact on throughput arising from the update step.

Building on this, we vary the amount of worker threads to show how \approachshort{} scales to fit available compute resources on a device.
This is an important aspect for (automated) allocation of compute resources in an intelligent dataplane---particularly when cohabiting with other dataplane programs---and has effects on ahead-of-time work scheduling which we examine later.
This also demonstrates the number of cores needed to achieve a given latency/throughput bound on a policy of representative complexity.
Moreover, to demonstrate how these costs vary as policy complexity increases, we vary the number of total dimensions included in the tiling.

@@

\newlength{\resultplotwidth}
\setlength{\resultplotwidth}{0.88\linewidth}

@@

\begin{figure}[t]
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core}
	}
	\caption{
		\Coopfw{}'s online learning performance improves with additional cores, on max size tasks (\num{129} work items). This requires \num{8} workers to offer greater online throughput than single-threaded in-NIC RL. Sharper performance increases occur when a new physical core is added (\numrange{7}{8}) or the scheduler works around a bottleneck (\numrange{13}{14}).\label{fig:vary-core}}
\end{figure}

\begin{figure}
	\resizebox{\resultplotwidth}{!}{
		\includegraphics{../plots/build/rl-perf-tester/vary-core-latency}
	}
	\caption{\Coopfw{}'s action latencies similarly improve with more cores. This requires \num{3} workers (4 total contexts) to improve upon the state-action latency of single-threaded in-NIC RL.\label{fig:vary-core-latency}}
\end{figure}

@@
